\research{Worksheet 2}{07-10-22}{Worksheet}

\begin{enumerate}
    \item \textbf{VAT at 17.5\% is added to a telephone bill of £80. What is the total amount to be paid?}
    \[80\times 1.175 = 94\]
    \item \textbf{A sweater is bought in a sale for £20 after a 15\% reduction. What was the original price of the sweater?}
    \begin{align*}
        \pounds20 &= 85\% \\
        \therefore \pounds0.253 &= 1\%\\
        \therefore \pounds23.50 &= 100\%        
    \end{align*}
    \item \textbf{A flat is bought for £180000. Each year the price rises by 5\%. How much is the flat worth after 3 years?}
    \[ 180000 \times10.5^3 = \pounds208372.50 \]
    \item \textbf{John bought a sports car for £18500. Two years later, he sold the car for £14000. Work out the percentage loss.}
    \begin{align*}
        \frac{\mathrm{change}}{\mathrm{original}}\times 100 &= \mathrm{percent\ diff}\\
        \frac{18500 - 14000}{18500}\times 100 &= 24.3\%
    \end{align*}
    \item \textbf{A car was bought for £12000. Each year it depreciated in value by 5\%. Work out the value of the car after 3 years.}
    \[12000\times0.95^3 = \pounds10288.50\]
    \item \textbf{On April 1st a train journey increases in price by 5\%. Six months later it increases in prce again by 2\%. Work out the final price of a ticket that originally cost £40.}
    \begin{align*}
        40\times 1.05 &= 42\\
        42 \times 1.02 &= \pounds 42.84
    \end{align*}
    \item \textbf{Toothpaste is sold in three different sized tubes.
    \begin{itemize}
        \item 50ml = £1.24,
        \item 75ml = £1.96,
        \item 100ml = £2.42.
    \end{itemize}
    Which of the tubes of toothpaste gives best value for money?}
    \begin{align*}
        50ml &= 0.0248\\
        75ml &= 0.02613\\
        100ml &= 0.0242 \leftarrow \mathrm{this\ is\ cheapest}
    \end{align*}
    \item \textbf{Express the following percentages as fractions:
    \begin{enumerate}
        \item 50\%
        \item 75\%
        \item 36\%
        \item 100\%
        \item 12.5\%
    \end{enumerate}}
    \begin{enumerate}
        \item $\displaystyle \frac{1}{2}$
        \item $\displaystyle \frac{3}{4}$
        \item $\displaystyle \frac{9}{25}$
        \item $\displaystyle \frac{1}{1} = 1$
        \item $\displaystyle \frac{1}{8}$
    \end{enumerate}
    \item \textbf{A photograph of length 27cm is to be enlarger in the ratio 3:7. What is the length of the enlarger photograph?}\\
    \textcolor{red}{...}
    \item \textbf{A recipe for 12 people uses 500g of plain flour. How much flour is needed for 18 people?}
    \[500\times1.5 = 750g\]
    \item \textbf{If 15 oranges costs £1.80, how much will 23 identical oranges cost?}
    \begin{align*}
        1.8\div 15 &=0.12\\
        0.12\times23 &= \pounds2.76
    \end{align*}
    \item \textbf{It took 4 people 7 days to build a fence. At the same rate how many people do we need if we want to build the fence in just 2 day?}
    \begin{align*}
        4\times 7 &=28 \mathrm{pepoles\ worth\ of\ work}\\
        28\div2&=14 \mathrm{people}
    \end{align*}
    \item \textbf{An inheritance is divided between three people in the ratio 4:9:2. If the lease amount received is £2300, calculate how much the other two people received.}
    \begin{align*}
        2300=2&\therefore 1=1150\\
        1150\times 4 &= 4600 (4)\\
        1150\times9 &= 10350 (9)
    \end{align*}
    \item \textbf{Make is taking the NetFun unit at the School of Computing. His mark for the coursework was 26 (50\% of overall mark). Make wants to have the final mark at least 70. Figure out whether it is l possible and if yes, calculate his lowest possible mark for the exam (50\% of overall mark). Answer the same questions for the final mark at least 60.}\\
    \textcolor{red}{...}
    \item \textbf{Tony walks 24km in 6 hours. (a) What is his average speed in km/h? (b) If he had taken 2 hours longer, what would have been his average speed?}
    \[(a)\ 4km/h\ \  (b)\ 3km/h\]
    \item \textbf{A marathon race was completed by 5 participants in the times given below: 2.7 hr, 8.3 hr, 3.5 hr, 5.1 hr, 4.9 hr. What is the average race time for this marathon?}
    \[ \frac{1.7 + 8.3 + 3.5 + 5.1 + 4.9}{5} = 4.9\]
    \item \textbf{[*] Two cyclists, initially 30 miles apart, travel towards each other. Cyclist A goes at 14 mph, and cyclist B at 16 mph. A fly flies back and forth between their noses at 30 mph. How far does the fly fly?}\\
    \textcolor{red}{...}
    \item \textbf{[*] Driving through roadworks on a motorway, the sign said 'speed limit 50; average speed calculated'. I noticed that for 12 minutes I was going at 60 mph. For how long do I have to go at 30 to be legal?}\\
    \textcolor{red}{...}
    \item \textbf{A painter wants to draw a big rectangle with sides 3 and 5 metres. He wants to use 3 different colours: blue, green, and yellow. A part with blue colour will cover 30\% of the total area of rectangle. The rest area will be painted equally by green and yellow colour. The cost of blue colour is £2.20 (per square metre), green colour is £2.50 (per square metre), yellow colour is £3 (per square metre). How much the painter will pay for the colours?}
    \begin{align*}
        3\times5 &= 15m^2\\
        15 \div 100 &= 0.15\\
        0.15 \times 30 &= 4.5\\
        4.5 \times 2.20 &= \pounds 9.90\\ \\
        0.15 \times 35 &= 5.25\\
        5.25 \times 2.5 &= \pounds 13.13\\
        5.25 \times 3 &= \pounds 15.75\\ \\
        9.9 + 13.13 + 15.75 &= \pounds 38.78
    \end{align*}
    \item \textbf{A die is thrown. Find the probability of
    \begin{enumerate}
        \item obtaining a score less than 6
        \item obtaining a score more than 6
        \item obtaining an even score less than 5
        \item obtaining an even score less than 2
    \end{enumerate}}
    \begin{enumerate}
        \item $\displaystyle \frac{5}{6}$
        \item $\displaystyle \frac{0}{6} = 0$
        \item $\displaystyle \frac{2}{6} = \frac{1}{3}$
        \item $\displaystyle \frac{0}{6} = 0$
    \end{enumerate}
    \item \textbf{A box contians 16 red blocks, 20 blue blocks, 24 orange blocks and 10 black blocks. A block is picked at random. Calculate the probability that the block is
    \begin{enumerate}
        \item black
        \item orange
        \item blue or red
        \item green
        \item not orange
    \end{enumerate}}
    $16+20+24+10 = 70$
    \begin{enumerate}
        \item $\displaystyle \frac{10}{70} = \frac{1}{7}$
        \item $\displaystyle \frac{24}{70}$
        \item $\displaystyle \frac{20+16}{70} = \frac{36}{70}$
        \item $\displaystyle \frac{0}{70} = 0$
        \item $\displaystyle \frac{70-24}{70} = \frac{46}{70}$
    \end{enumerate}
\end{enumerate}