\lecture{Introduction To Module}{27-09-2022}{10:00}{Zhaojie}{Zoom}

The goal of this module is to: help with maths involved in programming; prepare for other technical units (including: databases, functional programming; discrete maths; theoretical computer science; computer vision; and more); and to gain confidence.

The maths element (this element) is 30\% of the Architecture and Operating Systems module. The two components are independent until the final grade for the module is produced.

\section*{Assessments}
There are two components to the assessments for this element of the module.
\subsection*{Online Tests}
Throughout the year, there are 7 online tests which will be completed through the Pearson MyMathsLab. Overall, these tests will equal 15\% of the overall module score. Each test should last for 20 minutes and has a practice test available which can be accessed anytime in the year, and as many times as you wish. The real test can only be attempted once. After you submit the test, you will get an instant score.

Calculators are permitted, however it may not always be advisable to use one.

\subsection*{End Of Year Test}
At the end of the year, there will be an end of year test. This is equal to 15\% of the overall module score.

\section*{Staff \& Support}
Jhaojie is the main lecturer for the module. He will be assisted by Bryan.

Outside of class, Xia and Kirsten are available from the academic tutors office to help where needed. They can be booked through Moodle.

A MathsCafe runs during term-time for drop in Maths support. It is held at the following times
\begin{itemize}
    \item Monday 12:00-14:00, LG learning and teaching space;
    \item Tuesday 12:00-14:00, LG learning and teaching space;
    \item Wednesday, 12:00-14:00, Zoom;
    \item Wednesday 14:00-16:00, Library 0.36;
    \item Thursday 12:00-14:00, LG learning and teaching space;
    \item Friday 12:00-14:00, LG learning and teaching space.
\end{itemize}


When writing working out for questions, full steps should be written down. This allows errors to be found and corrected.

Take a pen and paper to practical classes.