\lecture{PRACTICAL: further joins}{01-12-22}{14:00}{Mark etc}{FTC 3}

\section*{Tutor Led}
We need to insert two more roles into the \texttt{Role} table.
\begin{sql}
INSERT INTO ROLE (role_name)
VALUES ('Cleaner');

INSERT INTO ROLE (role_name)
VALUES ('Pre Sales');
\end{sql}
Then run the following.
\begin{sql}
SELECT count(*)
FROM ROLE;
\end{sql}
This generates the following output
\begin{pseudo*}
 count 
-------
     7
(1 row)
\end{pseudo*}

\section*{Student Tasks}
1. Write a query that correctly displays the staff members first and last names, their email addresses and their roles. Use the method that uses the \texttt{JOIN} keyword. Copy the code and answer below.
\begin{sql}
SELECT staff.staff_fname, staff.staff_lname, staff.home_email, role.role_name FROM staff
JOIN role on staff.role = role.role_id;
\end{sql}
\begin{pseudo*}
 staff_fname | staff_lname |         home_email          |    role_name    
-------------+-------------+-----------------------------+-----------------
 Montgomery  | Housegoe    | mhousegoe2@ucoz.ru          | Order Picker
 Niel        | Welsby      | nwelsby0@rambler.ru         | Final Packer
 Jillene     | Revitt      | jrevitt8@cornell.edu        | Post Sales
 Harriette   | Fewster     | hfewster7@independent.co.uk | Post Sales
 Aura        | Clewlowe    | aclewlowe5@google.com.au    | Post Sales
 Hanan       | Gloster     | hgloster3@blogger.com       | Customer Retain
 Nikoletta   | Shrimpton   | nshrimpton1@unblog.fr       | Customer Retain
 Tim         | Illem       | tillem9@dedecms.com         | Misc
 Nell        | Olsson      | nolsson6@jiathis.com        | Misc
 Janeva      | Gillicuddy  | jgillicuddy4@altervista.org | Misc
(10 rows)
\end{pseudo*}
2. Rewrite the query created in 1 but this time use the \texttt{WHERE} keyword. Copy the code and answer below.
\begin{sql}
SELECT staff.staff_fname, staff.staff_lname, staff.home_email, role.role_name FROM staff, role
WHERE staff.role = role.role_id;
\end{sql}
\begin{pseudo*}
 staff_fname | staff_lname |         home_email          |    role_name
-------------+-------------+-----------------------------+-----------------
 Montgomery  | Housegoe    | mhousegoe2@ucoz.ru          | Order Picker
 Niel        | Welsby      | nwelsby0@rambler.ru         | Final Packer
 Jillene     | Revitt      | jrevitt8@cornell.edu        | Post Sales
 Harriette   | Fewster     | hfewster7@independent.co.uk | Post Sales
 Aura        | Clewlowe    | aclewlowe5@google.com.au    | Post Sales
 Hanan       | Gloster     | hgloster3@blogger.com       | Customer Retain
 Nikoletta   | Shrimpton   | nshrimpton1@unblog.fr       | Customer Retain
 Tim         | Illem       | tillem9@dedecms.com         | Misc
 Nell        | Olsson      | nolsson6@jiathis.com        | Misc
 Janeva      | Gillicuddy  | jgillicuddy4@altervista.org | Misc
(10 rows)
\end{pseudo*}

3. List the customer first and last names with their email addresses and the product names of the products they have ordered. But only for the customers who live in Waekolong. Copy the code and the answer below.
\begin{sql}
SELECT customer.cust_fname, customer.cust_lname, customer.email, product.prod_name FROM customer
JOIN cust_order ON customer.cust_id=cust_order.cust_id
JOIN manifest ON cust_order.cust_ord_id=manifest.cust_ord_id
JOIN product on manifest.prod_id=product.prod_id
WHERE customer.town='Waekolong';
\end{sql}
\begin{pseudo*}
   cust_fname    | cust_lname |          email          |                   prod_name
-----------------+------------+-------------------------+------------------------------------------------        
 Marie-françoise | Currier    | acurrier0@economist.com | Vision-oriented attitude-oriented core
 Marie-françoise | Currier    | acurrier0@economist.com | Balanced client-server product
 Marie-françoise | Currier    | acurrier0@economist.com | Exclusive client-server array
 Marie-françoise | Currier    | acurrier0@economist.com | Universal encompassing conglomeration
 Marie-françoise | Currier    | acurrier0@economist.com | Synergistic homogeneous ability
 Marie-françoise | Currier    | acurrier0@economist.com | Universal exuding protocol
 Marie-françoise | Currier    | acurrier0@economist.com | Universal global hub
 Marie-françoise | Currier    | acurrier0@economist.com | Balanced real-time info-mediaries
 Marie-françoise | Currier    | acurrier0@economist.com | Integrated 24/7 interface
 Marie-françoise | Currier    | acurrier0@economist.com | Re-engineered explicit software
 Marie-françoise | Currier    | acurrier0@economist.com | Customizable cohesive capacity
 Marie-françoise | Currier    | acurrier0@economist.com | Robust mission-critical complexity
 Marie-françoise | Currier    | acurrier0@economist.com | Organic clear-thinking system engine
 Marie-françoise | Currier    | acurrier0@economist.com | Stand-alone composite Graphical User Interface
(14 rows)
\end{pseudo*}

4. Write a query that returns all categories and the product names and order the output into category order. Copy the code and the answer below.
\begin{sql}
SELECT category.cat_name, product.prod_name FROM category
JOIN product ON product.prod_cat = category.cat_id
ORDER BY category.cat_name;
\end{sql}
\begin{pseudo*}
  cat_name   |                    prod_name
-------------+--------------------------------------------------
 Health      | Exclusive multimedia middleware
 Health      | Pre-emptive holistic intranet
 Health      | Ameliorated next generation orchestration
 Health      | Monitored asynchronous function
 Health      | Right-sized mission-critical pricing structure
 Health      | Profound human-resource forecast
 Health      | Realigned client-driven database
 Health      | Seamless optimal leverage
 Health      | User-friendly encompassing array
 Health      | Customizable cohesive capacity
...
(100 rows)
\end{pseudo*}

5. Rewrite the query for Q4 so that the output is ordered by category, then the product id. Copy the code and the answer below.
\begin{sql}
SELECT category.cat_name, product.prod_name FROM category
JOIN product ON product.prod_cat = category.cat_id
ORDER BY category.cat_name, product.prod_id;
\end{sql}
\begin{pseudo*}
  cat_name   |                    prod_name
-------------+--------------------------------------------------
 Health      | Balanced client-server product
 Health      | Pre-emptive holistic intranet
 Health      | Multi-tiered explicit paradigm
 Health      | Monitored asynchronous function
 Health      | Right-sized mission-critical pricing structure
 Health      | Open-architected homogeneous concept
 Health      | Fully-configurable full-range interface
 Health      | Customizable cohesive capacity
 Health      | Seamless optimal leverage
 Health      | Realigned client-driven database
...
(100 rows)
\end{pseudo*}

6. How can you prove that the product id is being used to do the ordering? (You may have already done this in Q5). Copy the code and the answer below.
\begin{sql}
SELECT category.cat_name, product.prod_name, product.prod_id FROM category
JOIN product ON product.prod_cat = category.cat_id
ORDER BY category.cat_name, product.prod_id;
\end{sql}
\begin{pseudo*}
  cat_name   |                    prod_name                     | prod_id 
-------------+--------------------------------------------------+---------
 Health      | Balanced client-server product                   |       4
 Health      | Pre-emptive holistic intranet                    |       6
 Health      | Multi-tiered explicit paradigm                   |      10
 Health      | Monitored asynchronous function                  |      20
 Health      | Right-sized mission-critical pricing structure   |      23
 Health      | Open-architected homogeneous concept             |      37
 Health      | Fully-configurable full-range interface          |      46
 Health      | Customizable cohesive capacity                   |      54
 Health      | Seamless optimal leverage                        |      57
 Health      | Realigned client-driven database                 |      59
...
(100 rows)
\end{pseudo*}

7. Write a query that will list all staff members first and last names along with their email addresses that are cleaners. Copy the code and the answer below.
\begin{sql}
SELECT staff.staff_fname, staff.staff_lname, staff.work_email FROM staff
JOIN role ON staff.role=role.role_id
WHERE role.role_name='Cleaner';
\end{sql}
\begin{pseudo*}
 staff_fname | staff_lname | work_email
-------------+-------------+------------
(0 rows)
\end{pseudo*}

8. How many staff are there who have the role Misc? Copy the code and the answer below.
\begin{sql}
SELECT count(*) FROM staff
JOIN role ON staff.role = role.role_id
WHERE role.role_name='Misc';
\end{sql}
\begin{pseudo*}
 count
-------
     3
(1 row)
\end{pseudo*}

9. What are the addresses of the staff that are returned by the query for Q8? You should output their first and last names too. Copy the code and the answer below.
\begin{sql}
SELECT staff.staff_fname, staff.staff_lname, concat_ws(' ', addr1, addr2, town, postcode) AS "address"
FROM staff
JOIN role ON role.role_id = staff.role
WHERE role.role_name='Misc';
\end{sql}
\begin{pseudo*}
 staff_fname | staff_lname |                    address
-------------+-------------+-----------------------------------------------
 Janeva      | Gillicuddy  | 6999 Kings Park Sachtjen Portsmouth PO05 5SF
 Nell        | Olsson      | 18424 Kenwood Court Farmco Havant PO22 6DL
 Tim         | Illem       | 85 Lillian Way Farragut Southsea PO93 0CN
(3 rows)
\end{pseudo*}

10. List the product id numbers with their names that start with the letters Re . Copy the code and the answer below.
\begin{sql}
SELECT prod_id, prod_name FROM product
WHERE prod_name LIKE 'Re%';
\end{sql}
\begin{pseudo*}
 prod_id |                    prod_name
---------+--------------------------------------------------
       7 | Re-engineered cohesive methodology
      11 | Re-engineered explicit software
      18 | Re-engineered actuating capability
      26 | Realigned 5th generation artificial intelligence
      39 | Realigned homogeneous hub
      56 | Reduced fresh-thinking process improvement
      59 | Realigned client-driven database
      76 | Re-engineered 24/7 knowledge base
(8 rows)
\end{pseudo*}

11. List the product id numbers with their names that have the word \texttt{value} in the name somewhere. Copy the code and the answer below.
\begin{sql}
SELECT prod_id, prod_name FROM product
WHERE prod_name LIKE '%value%';
\end{sql}
\begin{pseudo*}
 prod_id |           prod_name
---------+-------------------------------
      80 | Profound value-added intranet
(1 row)
\end{pseudo*}

12. List the product names along with their id numbers that have \texttt{Value} somewhere in their name. Copy the code and the answer below
\begin{sql}
SELECT prod_id, prod_name FROM product
WHERE prod_name LIKE '%Value%';
\end{sql}
\begin{pseudo*}
 prod_id | prod_name
---------+-----------
(0 rows)
\end{pseudo*}

13. List the customer first and last names along with their email addresses, the customer order id, the category names and the product names for orders that have been placed for all products that have the word \texttt{able} in the name. (The case matters). Order by the category and the product name. The output should have the category names in alphabetical order then within each category the products should be ordered in alphabetical order. Copy the code and the answer below.
\begin{sql}
SELECT customer.cust_fname, customer.cust_lname, customer.email, cust_order.cust_ord_id, category.cat_name, product.prod_name from customer
JOIN cust_order ON customer.cust_id=cust_order.cust_id
JOIN manifest ON cust_order.cust_ord_id=manifest.cust_ord_id
JOIN product on manifest.prod_id=product.prod_id
JOIN category on category.cat_id=product.prod_cat
WHERE product.prod_name LIKE '%able%'
ORDER BY category.cat_name, product.prod_name;
\end{sql}
\begin{pseudo*}
   cust_fname    |    cust_lname    |              email              | cust_ord_id |  cat_name  |               
 prod_name
-----------------+------------------+---------------------------------+-------------+------------+-----------------------------------------
 Bérengère       | Menendez         | amenendez3@dell.com             |          64 | Health     | Customizable cohesive capacity
 Marie-françoise | Currier          | acurrier0@economist.com         |         133 | Health     | Customizable cohesive capacity
 Bérengère       | Menendez         | amenendez3@dell.com             |         102 | Health     | Fully-configurable full-range interface
 Chadd           | Franz-Schoninger | cfranzschoninger3@google.com.hk |           7 | Health     | Team-oriented stable project
 Chadd           | Franz-Schoninger | cfranzschoninger3@google.com.hk |          81 | Health     | Team-oriented stable project
 Bénédicte       | Dozdill          | cdozdill1@amazon.de             |          24 | Kid's Wear | Configurable analyzing solution
 Bérengère       | Menendez         | amenendez3@dell.com             |          21 | Kid's Wear | Configurable analyzing solution
 Bérengère       | Menendez         | amenendez3@dell.com             |         113 | Kid's Wear | Configurable analyzing solution
 Jobey           | Boeter           | jboeter0@mail.ru                |          91 | Kid's Wear | Configurable analyzing solution
 Jobey           | Boeter           | jboeter0@mail.ru                |          39 | Outdoor    | Switchable tangible product
 Jobey           | Boeter           | jboeter0@mail.ru                |          26 | Outdoor    | Switchable tangible product
 Vikky           | Eke              | veke4@elegantthemes.com         |         105 | Sport      | Configurable methodical firmware
 Vikky           | Eke              | veke4@elegantthemes.com         |         118 | Sport      | Customizable well-modulated encryption
 Pélagie         | Hachard          | fhachard4@blinklist.com         |          89 | Sport      | Virtual stable Graphic Interface
(14 rows)
\end{pseudo*}