\lecture{Introduction to Module}{26-09-22}{16:00}{Farzad}{RB LT1}

\section*{Division of the Module}
This module is split into two parts: computer (this part) which is worth 70\% and maths (the other part) which is worth 30\%. The two parts are run completely independently of each other. The only time they come together is when the final overall score is calculated. 

There are two separate Moodle pages (one for Computer and one for Maths)

\section*{Computer Module assessments}
For the Computer section of the module, there are two assessments. One is in January 2023, which will be a Computer Based Test (covering content taught in the first teaching block). It is worth 30\% of the over module score. The second is in the May/June 2023 assessment period. It will be computer based. This assessment will be worth 40\% of the overall module score.

Both assessments are closed book however a formula sheet will be provided for the January assessment. Nothing is provided for the May/June assessment.

The pass mark for the entire module is 40\%, this score is generated from all the computer assessments AND all the maths assessments.

\section*{Module structure}
There will be a one hour lecture per week, where content is introduced to us. This will be delivered using worksheets for the first 10 weeks.

There will also a practical session each week where the cohort is split into smaller groups. These sessions will be a chance to practice the ideas introduced in the lectures. There will be more members of staff around at the practical sessions to help out.
\begin{link}{More Information on Practical Sessions}
There are practical session guidelines available in the induction slides or on Moodle.
\end{link}

\begin{link}{Content in each Week}
There is a teaching plan on Moodle which outlines the content covered each week as well as the weeks in which the exams will be held.
\end{link}