\lecture{Introduction To Module}{27-09-22}{13:00}{Nadim}{RB LT1}

This module is split into four items. The first of these is a video coursework project. The second, third and fourth are combined into an end of year exam. Each item is worth 25\% of the overall module grade, therefore the end of year exam is worth 75\%.

This module was created because the University doesn't do modules which are smaller than 20 credits and none of the items are big enough to be their owen module.

To pass the module, you need to score at least 40\% overall, not in each individual item.

The four items are taught by different lecturers and are shown below
\begin{enumerate}
    \item Video coursework
    \item Web
    \item Security
    \item Either UXD or DB (which we do is decided for us)
\end{enumerate}

Each item will be introduced to us when we start that item.

\section*{Item 1: Coursework}
This is due at 11pm on 16th December 2022. It is to be uploaded to YouTube, with a link to the video put in a PDF document which is uploaded to Moodle. 

This item is able to be done either in groups or by individuals, it should be very easy to get a good grade in it.

The task is to select a conversation and analyse the conversation, using supporting research and references. 

The video should be at most 4 minutes long, it can be a mixture of different takes edited together. There are a number of different pieces of editing software available on AppsAnywhere. 

\begin{link}{More Information}
More information for this coursework can be found on a Google Doc which is linked from the Moodle page. This Google Doc links to the official University Coursework information document, the conversations and outlines the lecture plan for the first half of TB1.
\end{link}

There is a one lecture per week and an optional drop in session per week. The drop in sessions are primarily there to answer quick questions.