\lecture{Matrices}{2023-01-31}{10:00}{Zhaojie}{Zoom}

A matrix is a set of numbers/ variables arranged in the form of a rectangle and enclosed in curved brackets. For example
\begin{equation*}
A = 
\begin{pmatrix}
    3 & -2 & 2 & 9\\
    2 & -1 & 0 & 0\\
    1 & 2 & 1 & 5
\end{pmatrix}
\mathrm{or\ }
B=
\begin{pmatrix}
    3\\
    2\\
    1\\
    -3
\end{pmatrix}
\end{equation*}

Each entry in a matrix is known as an element. The size of a matrix is given by its number of rows $\times$ number of columns (in that order), for example the size of $A$ is $3 \times 4$ and the size of $B$ is $4 \times 1$.

The elements of a matrix can be referred to by their row and column number. The elements of the matrix
\[A= \begin{pmatrix}
    3 & -2\\
    2 & -1\\
    1 & 2
\end{pmatrix} \]
can be referred using elements
\[\begin{pmatrix}
    a_{11} & a_{12}\\
    a_{21} & a_{22}\\
    a_{31} & a_{32}
\end{pmatrix}\]
The main diagonal of a matrix ($A$) is the collection of elements $a_{ij}$ where $i=j$

\section*{Addition \& Subtraction of Matrices}
Certain combinations of matrices can be added, subtracted and multiplied (but never divided). If two matrices have the same size, they can be added or subtracted by simply adding or subtracting the corresponding elements. Examples of addition and subtraction can be seen below.
\begin{equation*}
\begin{pmatrix}
    3 & -2\\
    2 & -1\\
    1 & 2
\end{pmatrix}
+
\begin{pmatrix}
    4 & -5\\
    -2 & 1\\
    0 & 2
\end{pmatrix}
=
\begin{pmatrix}
    3+4 & -2-5\\
    2-2 & -1+1\\
    1+0 & 2+2
\end{pmatrix}
=
\begin{pmatrix}
    7 & -7\\
    0 & 0\\
    1 & 4
\end{pmatrix}
\end{equation*}

\begin{equation*}
\begin{pmatrix}
    3 & -2\\
    2 & -1
\end{pmatrix}
-
\begin{pmatrix}
    4 & -5\\
    -2 & 1
\end{pmatrix}
=
\begin{pmatrix}
    3-4 & -2+5\\
    2+2 & -1-1
\end{pmatrix}
= 
\begin{pmatrix}
    -1 & 3\\
    4 & -2
\end{pmatrix}
\end{equation*}

\section*{Multiplication of Matrices}
Two matrices can be multiplied together if the number of columns in the first matrix is the same as the number of rows in the second matrix. 

If $A$ has size $p \times q$ and $B$ has size $q \times s$ then $AB$ has size $p \times s$. 

The element in row $i$ and column $j$ of $AB$ is the result of the `multiplication' of the row $i$ by column $j$, for example
\[
A= 
\begin{pmatrix}
    3 & 2\\
    2 & -1\\
    -2 & 1
\end{pmatrix}
,\quad
B= 
\begin{pmatrix}
    4 & 1 & -1 & 0\\
    1 & -1 & 1 & -2
\end{pmatrix}
\]
Then the element in row 1 and column 1 of $AB$ is
\[ [3,2] \times [4,1] = 3 \times 4 + 2 \times 1 = 14 \]
The element in row 2 and column 3 of $AB$ is
\[ [2, -1] \times [-1, 1] = 2 \times (-1) + (-1) \times 1 = -3 \]

To multiply a matrix by a number, multiply each element by that number individually.